A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-02-08 response/response.tex

\documentclass{article}
\usepackage[left=1.0in,right=1.0in,top=1.0in,bottom=1.0in]{geometry}
\usepackage{fancyhdr}
\usepackage{color}
\usepackage{amsmath}
\definecolor{light-gray}{gray}{0.5}

\begin{document}

\begin{center}
    {\LARGE \bf Point-by-Point Response to Reviewer's Comments} \\
     Author's Responses in {\color{red} Red}
\end{center}

%%%% First Reviewer %%%%
\section{Information to the authors}
%
\subsection*{Larger issues}
%
\begin{enumerate}
    \item
         Some parts of the manuscript suggest that the authors introduced a new method,
         i.e., the complex-step (CS) method. For example, in the introduction, they say
         “A goal of this study was to develop and evaluate a new, accurate, and practical
         method ...” Another example appears in page 3, where the authors say “The aim
         of this paper is not only to introduce the new complex-step method in the context
         of evaluating tangent-stiffness matrices ..”. However, the CS method has already
         appeared in many publications since 1998 [1]. Please revise the way you refer to
         the CS method along the manuscript

{\color{red}
  \begin{itemize}
    \item We have added a paragraph at the end of Section 1 that highlights the difference between peridynamics and previous Òparticle methods.Ó The main difference is that in peridynamics, we numerically approximate a different set of continuum equations.
  \end{itemize}}

### 2014-02-08 jacobian_calc_CMAME.tex
response to I.1 started

\begin{abstract} %% Text of abstract Shown is a retrospective comparative study of tangent-stiffness
In order to maintain the quadratic convergence properties of Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. A goal of the study described in this paper was to establish the suitability of an under-explored method for numerical computation of tangent-stiffness operators, referred to as ``complex-step'', and compare the new method with other techniques for numerical derivative calculation: automatic differentiation, forward finite-difference, and central finite-difference. The complex-step method was newly implemented in a massively parallel computational peridynamics code for the purpose of this comparison. The methods were compared through in situ profiling of the code for accuracy, speed, efficiency, and parallel scalability. The research provides data that can serve as practical guide for code developers and analysts faced with choosing which method best suits the needs of their application code. Additionally, motivated by the reproducible research movement, all the of the code, examples, and workflow to regenerate the data and figures in this paper are provided as open source.

@@

\section{Introduction}
%\subsection{Motivation}
\label{sec:intro} 
In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

A goal of this study was to develop and evaluate a new, accurate, and practical method for calculating tangent-stiffness matrices against established methods.  This new method is based on a complex number Taylor series expansion and referred to as CTSE or the ``complex-step'' method.  The
distinction of ``accurate'' is defined by comparison of the new method to popular finite differencing techniques used for computing tangent-stiffness matrices and with the exact to machine precision algorithmically consistent derivatives computed with \emph{automatic differentiation}.

@@

The scope of the application component of the study was limited to a single material model, an elastic peridynamic solid, implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the new method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of finite-difference, central-difference, and automatic-differentiation methods needed for comparison to the new complex-step method implementation, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters. Agile-components is a software development term meaning to structure a software project for continuous development to respond to evolving specifications by making much use of techniques from object-oriented programming and reusing proven pre-existing code to reduce development time and increase reliability. The modularity of \emph{Peridigm} allowed the adding of new features, such as required by the study.

The aim of this paper is not only to introduce the new complex-step method in the context of evaluating tangent-stiffness matrices, but to serve as a review for other differentiation techniques useful for solving non-linear systems with Newton's method. After presenting background information on the underlying methods to be discussed, presented in this paper are: A) a description of tangent-stiffness matrices, B) detailed directions for producing tangent-stiffness matrices with each the methods identified above, C) a description and justification of the new complex-step method for calculating tangent-stiffness matrices, D) a description of implementing complex-step in the \emph{Peridigm} software, E) a description of the quantities-of-interest used to rank the methods, F) a presentation and analysis of the results of the comparative study, and G) conclusions and thoughts on potential future work. Finally, in the interest of replicable research, the reader is referred to the corresponding author's website for \emph{Peridigm} C++ source-code and data necessary to reproduce the results presented here. Additionally, included on the website is an example serial C++ library with classes for solving non-linear systems using the techniques discussed here, as well as two validation and verification example problems which show how to use the library. The purpose of the serial C++ library is to demonstrate in a less complex manner than in the \emph{Peridigm} code how one would use automatic-differentiation or complex-step to better encourage their use and discussion. This open source resource will be referred to as the \emph{paper repository} in what follows.

@@

\subsubsection{Quantities of interest} 
\label{JGAM} 
As stated in the Introduction, a goal of the study was to compare CS, FD, and CD on the basis of accuracy. AD was omitted from the comparison because it served as the standard of accuracy for the other methods in the absence of appropriate analytical forms for the tangent-stiffness matrix associated with the system solved in the study. The assumption that AD is accurate enough to serve as a standard is supported by ADs implementation as a computerized chain rule as explained in Section~\ref{ADsubsection}.

@@

\subsubsection{Data Reduction} 
The output produced by \emph{Perdigm} including the additional accuracy and speed measurements was redirected from the console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).  Directories and filenames were chosen so that data would be easy to sort by individual test run.  \emph{Python} scripts were then used to post-process and plot the data. These scripts averaged accuracy or speed data for a tests corresponding to a particular peridynamic node density over all iterations for that run, and then plotted this averaged data  as a function of peridynamic node density as it varied from test to test in the series.  Similarly, for the parallel tests, the data was averaged and plotted in much the same way, but as a function of number of compute cores used to solve the problem. Confidence intervals were not calculated on the measurements. The main reason is that network traffic from other users makes running tests on the HPC cluster a time varying process. These data reduction and plotting scripts can be found on the corresponding author's website.

@@

\subsection{Accuracy Data}
Data for accuracy measurements were taken according to the methods mentioned in Section~\ref{tcs}, and reduced using equation~(\ref{eqn:accuracy}). The averaged results of these accuracy measurements for the serial tests can be seen in Fig.~\ref{fig:serial_accuracy}. Averaged Frobenius norm (referred to as $l^2$-norm in the figure labels for conciseness) of the element-wise difference between a given TS matrix and the TS matrix produced by the AD method is plotted as a function of the number of nonzero TS matrix elements. CS was shown to be the most accurate of the methods when compared to AD and was roughly \emph{$6$ orders of magnitude more accurate than CD} and \emph{$9$ orders of magnitude more accurate than FD}. This accuracy ranking order was maintained for each test in the series. It should be pointed out that compared to the $l^2$-norms of any of the TS matrices by themselves, the magnitude of the accuracy metric calculated with Eqn.~\ref{eqn:accuracy} for any of the methods is relatively small. These values for the $l^2$-norms of the TS matrices appear in the simulation data which is available in the paper repository. The effect of these accuracy differences on convergence rate of the Newton solver was not determined.  
%
The accuracy of FD and CD could be improved by using a smaller probe distance, $h$; however, if $h$ is too small, then the methods can suffer from severe effects, including solution divergence, of the round-off error associated with subtracting two numbers that are very close to one another.  The value of $h$ used in this study is the default value used in \emph{Peridigm} which is heuristically derived based on computational node spacing.  One of the beauties of the CS method is that since dependence on $h$ decreases for small $h$, which a numerical example in \cite[Table 1]{squire1998using} demonstrates, $h$, cannot be too small, therefore it can be set to machine epsilon and the analysis performed without the issues associated with round-off error. The value chosen for $h$ for CS used in this study was $\num{1.E-100}$ to demonstrate this.
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_accuracy.pgf}}
  \caption{Serial test series accuracy measurements.}
  \label{fig:serial_accuracy}
\end{figure}
%
The averaged results of the accuracy measurements taken for the parallel test series can be seen in Fig.~\ref{fig:multi_accuracy}. The ordinate axis units remain unchanged from Fig.~\ref{fig:serial_accuracy}, but the abscissa now indicates number of compute cores used to solve the test problem. Accuracy trends matched those for the single core tests. 
%
\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_accuracy.pgf}}
  \caption{Multicore test series accuracy measurements.}
  \label{fig:multi_accuracy}
\end{figure}

@@

\section{Conclusions}
%
Based on the experimental results, it appears that for users of \emph{Peridigm} there is a case to be made for using AD for reasons of performance and accuracy. However, certain parallel tests done in this study showed a small performance gain over AD with the FD method.  While the results showed that CS produced more accurate tangent-stiffness matrices than CD and FD under the parameters of the tests, it was not determined whether or not this is a clear advantage of CS over FD in terms of accuracy of final predicted displacements and speed of convergence. The reader is reminded that the time required to complete an iteration must be considered along with the total number of iterations taken order to measure speed of convergence. The goal of the comparative study was to benchmark single iteration performance and not convergence rate. For new developments it is recommended that combined accuracy and speed of iteration measurements be conducted along with entirely separate convergence studies. It should be mentioned that a related work, \cite{perez2012numerical}, compares the speed and accuracy of complex-step to finite difference for producing tangent-stiffness matrices in solid mechanics simulations using complex inelastic material model, where the researchers similarly found CS to be accurate yet relatively computationally expensive compared to FD and CD. In that work, full convergence studies were also done and indicated that quadratic convergence could be achieved with complex-step all of the methods \cite[p.28]{perez2012numerical}.

The CS method was shown to be highly accurate in calculating tangent-stiffness matrices by the standards of this study, and it is slower than other methods for a naive implementation. In the context of future use of CS in \emph{Peridigm}, CS holds the advantage of relying on byte-copyable data types, while AD requires a complicated serialization and deserialization in order to function as byte-copyable. This is fine for copying in between MPI processes, but deserialization on an accelerator or GPU is very difficult, as this would require compiling a version of the AD library being used that were compatible with the accelerator or GPU being used. What this means is that if \emph{Peridigm} or any simulation software were to be reformulated to take advantage of accelerators or GPUs, the flexibility of complex-step owed to its simple implementation may make it more viable for that application given its high accuracy. Additional performance gains may be achieved through the use of the templated \emph{Tpetra} package in \emph{Trilinos} which would allow the use of complex data types natively, and alleviate the expense of the dynamic-casts that were utilized in this study.  As a final comment, it should be noted that from a code development perspective the CS method may be the easiest of the methods to implement; It only requires one functional evaluation, there is no heuristic choice of probe distance, and there is no dependence on an external AD library.  From this perspective, the method may still be an attractive method for TS matrix evaluation.

>>>>

\begin{abstract} %% Text of abstract Shown is a retrospective comparative study of tangent-stiffness

In order to maintain the quadratic convergence properties of Newton's method in
quasi-static nonlinear analysis of solid structures it is crucial to obtain
accurate, algorithmically consistent tangent-stiffness matrices. A goal of the
study described in this paper was to establish the suitability of an
under-explored method for numerical computation of tangent-stiffness operators,
referred to as ``complex-step'', and compare the \note[Michael]{re: I.1}
\remove[Michael]{new} method with other techniques for numerical derivative
calculation: automatic differentiation, forward finite-difference, and central
finite-difference. The complex-step method was \note[Michael]{re: I.1, could
suggest claim of novelty} \remove[Michael]{newly} implemented in a massively
parallel computational peridynamics code for the purpose of this comparison.
The methods were compared through in situ profiling of the code for accuracy,
speed, efficiency, and parallel scalability. The research provides data that
can serve as practical guide for code developers and analysts faced with
choosing which method best suits the needs of their application code.
Additionally, motivated by the reproducible research movement, all the of the
code, examples, and workflow to regenerate the data and figures in this paper
are provided as open source.

@@

\section{Introduction}
%\subsection{Motivation}
\label{sec:intro}

In order to maintain the quadratic convergence properties of the first-order
Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in
quasi-static nonlinear analysis of solid structures it is crucial to obtain
accurate, algorithmically consistent tangent-stiffness matrices. For an
extremely small class of nonlinear material models, these consistent
tangent-stiffness operators can be derived analytically; however, most often in
practice, they are found through numerical approximation of derivatives.

A goal of this study was to develop and evaluate a \note[Michael]{re: I.1}
\change[Michael]{new}{relatively underexplored}, accurate, and practical method
for calculating tangent-stiffness matrices against established methods.  This
new method is based on a complex number Taylor series expansion and referred to
as CTSE or the ``complex-step'' method.  The distinction of ``accurate'' is
defined by comparison of the \note[Michael]{re: I.1}\remove[Michael]{new} method to popular finite differencing
techniques used for computing tangent-stiffness matrices and with the exact to
machine precision algorithmically consistent derivatives computed with
\emph{automatic differentiation}.

@@

The scope of the application component of the study was limited to a single
material model, an elastic peridynamic solid, implemented in the computational
peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific
application provided a practical framework for implementing the
\note[Michael]{re: I.1}\remove[Michael]{new} method and solving engineering
problems to generate the data needed to compare the methods. In particular,
\emph{Peridigm} was chosen because it combined several helpful
characteristics: the utilization of Newton's method and tangent-stiffness
matrices for solving nonlinear quasi-static problems, prior inclusion of
finite-difference, central-difference, and automatic-differentiation
methods needed for comparison to the new complex-step method
implementation, and being an agile-components code that makes use of
distributed computing data structures via Trilinos \cite{trilinos} for
efficient parallelization on large clusters. Agile-components is a software
development term meaning to structure a software project for continuous
development to respond to evolving specifications by making much use of
techniques from object-oriented programming and reusing proven pre-existing
code to reduce development time and increase reliability. The modularity of
\emph{Peridigm} allowed the adding of new features, such as required by the
study.

The aim of this paper is not only to \note[Michael]{re: I.1}\change[Michael]{introduce the new}{study the} complex-step method \change[Michael]{in
the context of}{of} evaluating tangent-stiffness matrices, but to serve as a review
for other differentiation techniques useful for solving non-linear systems with
Newton's method. After presenting background information on the underlying
methods to be discussed, presented in this paper are: A) a description of
tangent-stiffness matrices, B) detailed directions for producing
tangent-stiffness matrices with each of the methods identified above, C) a
description and justification of the \note[Michael]{re: I.1}\remove[Michael]{new}complex-step method for calculating
tangent-stiffness matrices, D) a description of implementing complex-step in
the \emph{Peridigm} software, E) a description of the quantities-of-interest
used to rank the methods, F) a presentation and analysis of the results of the
comparative study, and G) conclusions and thoughts on potential future work.
Finally, in the interest of replicable research, the reader is referred to the
corresponding author's website for \emph{Peridigm} C++ source-code and data
necessary to reproduce the results presented here. Additionally, included on
the website is an example serial C++ library with classes for solving
non-linear systems using the techniques discussed here, as well as two
validation and verification example problems which show how to use the library.
The purpose of the serial C++ library is to demonstrate in a less complex
manner than in the \emph{Peridigm} code how one would use
automatic-differentiation or complex-step to better encourage their use and
discussion. This open source resource will be referred to as the \emph{paper
repository} in what follows.

@@

\subsubsection{Quantities of interest} \label{JGAM}

As stated in the Introduction, a goal of the study was to compare CS, FD, and
CD on the basis of accuracy. AD was omitted from the comparison because it
served as the standard of accuracy for the other methods in the absence of
appropriate analytical forms for the tangent-stiffness matrix associated with
the system solved in the study. The assumption that AD is accurate enough to
serve as a standard is supported by ADs implementation as a computerized chain
rule as explained in Section~\ref{ADsubsection}.

@@

\subsubsection{Data Reduction}

The output produced by \emph{Perdigm} including the additional accuracy and
speed measurements was redirected from the console to text files by the
resource manager, \emph{SLURM} (Simple Linux User Resource Manager).
Directories and filenames were chosen so that data would be easy to sort by
individual test run.  \emph{Python} scripts were then used to post-process and
plot the data. These scripts averaged accuracy or speed data for a tests
corresponding to a particular peridynamic node density over all iterations for
that run, and then plotted this averaged data  as a function of peridynamic
node density as it varied from test to test in the series.  Similarly, for the
parallel tests, the data was averaged and plotted in much the same way, but as
a function of number of compute cores used to solve the problem. Confidence
intervals were not calculated on the measurements. The main reason is that
network traffic from other users makes running tests on the HPC cluster a time
varying process. These data reduction and plotting scripts can be found on the
corresponding author's website.

@@

\subsection{Accuracy Data}

Data for accuracy measurements were taken according to the methods mentioned in
Section~\ref{tcs}, and reduced using equation~(\ref{eqn:accuracy}). The
averaged results of these accuracy measurements for the serial tests can be
seen in Fig.~\ref{fig:serial_accuracy}. Averaged Frobenius norm (referred to as
$l^2$-norm in the figure labels for conciseness) of the element-wise difference
between a given TS matrix and the TS matrix produced by the AD method is
plotted as a function of the number of nonzero TS matrix elements. CS was shown
to be the most accurate of the methods when compared to AD and was roughly
\emph{$6$ orders of magnitude more accurate than CD} and \emph{$9$ orders of
magnitude more accurate than FD}. This accuracy ranking order was maintained
for each test in the series. It should be pointed out that compared to the
$l^2$-norms of any of the TS matrices by themselves, the magnitude of the
accuracy metric calculated with Eqn.~\ref{eqn:accuracy} for any of the methods
is relatively small. These values for the $l^2$-norms of the TS matrices appear
in the simulation data which is available in the paper repository. The effect
of these accuracy differences on convergence rate of the Newton solver was not
determined.  
%
The accuracy of FD and CD could be improved by using a smaller probe distance,
$h$; however, if $h$ is too small, then the methods can suffer from severe
effects, including solution divergence, of the round-off error associated with
subtracting two numbers that are very close to one another.  The value of $h$
used in this study is the default value used in \emph{Peridigm} which is
heuristically derived based on computational node spacing.  One of the beauties
of the CS method is that since dependence on $h$ decreases for small $h$, which
a numerical example in \cite[Table 1]{squire1998using} demonstrates, $h$,
cannot be too small, therefore it can be set to machine epsilon and the
analysis performed without the issues associated with round-off error. The
value chosen for $h$ for CS used in this study was $\num{1.E-100}$ to
demonstrate this.
%
\begin{figure}[tbp] \centering
\scalebox{1.0}{\input{./figs/serial_accuracy.pgf}} \caption{Serial test series
accuracy measurements.} \label{fig:serial_accuracy} \end{figure}
%
The averaged results of the accuracy measurements taken for the parallel test
series can be seen in Fig.~\ref{fig:multi_accuracy}. The ordinate axis units
remain unchanged from Fig.~\ref{fig:serial_accuracy}, but the abscissa now
indicates number of compute cores used to solve the test problem. Accuracy
trends matched those for the single core tests. 
%
\begin{figure}[tbp] \centering
\scalebox{1.0}{\input{./figs/multi_accuracy.pgf}} \caption{Multicore test
series accuracy measurements.} \label{fig:multi_accuracy} \end{figure}

@@

\section{Conclusions}
%
Based on the experimental results, it appears that for users of \emph{Peridigm}
there is a case to be made for using AD for reasons of performance and
accuracy. However, certain parallel tests done in this study showed a small
performance gain over AD with the FD method.  While the results showed that CS
produced more accurate tangent-stiffness matrices than CD and FD under the
parameters of the tests, it was not determined whether or not this is a clear
advantage of CS over FD in terms of accuracy of final predicted displacements
and speed of convergence. The reader is reminded that the time required to
complete an iteration must be considered along with the total number of
iterations taken order to measure speed of convergence. The goal of the
comparative study was to benchmark single iteration performance and not
convergence rate. For new developments it is recommended that combined accuracy
and speed of iteration measurements be conducted along with entirely separate
convergence studies. It should be mentioned that a related work,
\cite{perez2012numerical}, compares the speed and accuracy of complex-step to
finite difference for producing tangent-stiffness matrices in solid mechanics
simulations using a complex inelastic material model, where the researchers
similarly found CS to be accurate yet relatively computationally expensive
compared to FD and CD. In that work, full convergence studies were also done
and indicated that quadratic convergence could be achieved with complex-step
\remove[Michael]{all of the methods} \cite[p.28]{perez2012numerical}.

The CS method was shown to be highly accurate in calculating tangent-stiffness
matrices by the standards of this study, and it is slower than other methods
for a naive implementation. In the context of future use of CS in
\emph{Peridigm}, CS holds the advantage of relying on byte-copyable data types,
while AD requires a complicated serialization and deserialization in order to
function as byte-copyable. This is fine for copying in between MPI processes,
but deserialization on an accelerator or GPU is very difficult, as this would
require compiling a version of the AD library being used that were compatible
with the accelerator or GPU being used. What this means is that if
\emph{Peridigm} or any simulation software were to be reformulated to take
advantage of accelerators or GPUs, the flexibility of complex-step owed to its
simple implementation may make it more viable for that application given its
high accuracy. Additional performance gains may be achieved through the use of
the templated \emph{TPetra} package in \emph{Trilinos} which would allow the
use of complex data types natively, and alleviate the expense of the
dynamic-casts that were utilized in this study.  As a final comment, it should
be noted that from a code development perspective the CS method may be the
easiest of the methods to implement; It only requires one functional
evaluation, there is no heuristic choice of probe distance, and there is no
dependence on an external AD library.  From this perspective, the method may
still be an attractive method for TS matrix evaluation.
